---
title: "Beta Distributed Graph Estimation"
author: "Eric Bridgeford"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Beta Graph Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  header-includes:
   - \usepackage{amsfonts}
   - \usepackage{amsmath}
   - \usepackage[linesnumbered,lined,boxed,commentsnumbered]{algorithm2e}
---

In this tutorial, we discuss our estimator of a beta distribution per edge for a given graph.

# Framework

## Setting

+ $\mathbb{G}: \Omega \rightarrow \mathcal{G}$ is a graph-valued RV with samples $G_i \sim \mathbb{G}$
+ For each $G_i \in \mathcal{G}$, we have $G_i = (V, E_i)$; that is, each $G_i$ is defined by a set of vertices $V$ and a set of edges $E_i$, where $w_i: V \times V \rightarrow \mathbb{R}$, and $e_{uv}^{(i)} \in \mathbb{R}$. That is, $w_i$ is the real-valued weight function for each graph.
+ $\mathbb{A}: \Omega \rightarrow \mathcal{A}$, a adjacency-matrix-valued RV with samples $A_i \sim \mathbb{A}$, where $\mathcal{A}$ is the space of possible adjacency-matrices and $A_i \in \mathcal{A}$.
+ $A_i \in \mathcal{A}$, and $\mathcal{A} \subseteq \mathbb{R}^{V \times V}$. 
+ Each graph $G_i$ can be represented as an adjacency-matrix $A_i$.

## Statistical Goal

Identify the sufficient parameters to characterize the distributions of each edge's weights.

## Model

Assume that the edge weights can be characterized by a beta RV; that is:

\begin{align}
  \mathbb{A}_{uv} \sim Beta(\alpha_{uv}, \beta_{uv})
\end{align}


# Estimators

## Beta Parameters

Using the method of moments, it is easy to see that:

\begin{align*}
  \mu_{uv} = \mathbb{E}_{\mathbb{A}_{uv}}[A_{uv}] = \frac{\alpha_{uv}}{\alpha_{uv} + \beta_{uv}} \\
  \sigma_{uv}^2 = \mathbb{E}_{\mathbb{A}_{uv}}[(A_{uv} - \mu_{uv})^2] = \frac{\alpha_{uv}\beta_{uv}}{(_{uv} + \beta_{uv})^2(\alpha_{uv} + \beta_{uv} + 1)}
\end{align*}

Where $\mu_{uv}$ is the sample mean and $\sigma_{uv}$ is the sample standard deviation. Solving this system of equations for $\alpha_{uv}$ and $\beta_{uv}$, we find that:

\begin{align*}
    \alpha_{uv} &= \left(\frac{1 - \mu_{uv}}{\sigma_{uv}^2} - \frac{1}{\mu_{uv}}\right)\mu_{uv}^2 \\
    \beta_{uv} &= \alpha_{uv} \left(\frac{1}{\mu_{uv}} - 1\right)
\end{align*}

# Pseudo Code

Below is bare-bones pseudo code for each method. Note that optional parameters are omitted.

```{r eval=FALSE}
# A function to estimate a beta distribution for each edge in a graph.
# Inputs
#   sample: a [n x m x p] element array, where we have p observations of nxm graphs.
# Outputs
#   alpha: a [n x m] matrix denoting the alpha parameter per edge.
#   beta: a [n x m] matrix denoting the beta parameter per edge.
beta_graph_estimator(sample):
  for u in 1:n:
    for v in 1:m:
      alpha[u, v], beta[u, v] = beta_est(sample[u, v, :])  # estimate the parameters given the u, v edge from all samples
  return alpha, beta

# a function to estimate a beta distribution for a sample.
# Inputs
#   sample: a [p] element vector that is between 0 and 1.
# Outputs
#   alpha: the alpha parameter of the method of moments estimate given the sample.
#   beta: the beta parameter of the method of moments estimate given the sample.
beta_est(sample):
  alpha = ((1 - mu)/sig^2 - 1/mu)*mu^2
  beta = alpha*(1/mu - 1)
  return alpha, beta
  
# a function to generate random beta-distributed samples.
# Inputs
#   alpha: a [n x m] matrix of the alpha parameters per edge.
#   beta: a [n x m] matrix of the beta parameters per edge.
#   p: the number of samples.
# Outputs
#   samp: a [n x m x p] array sampling from the [n x m] graph RV p times.
sample_beta_graph(alpha, beta, p):
  for (i in 1:n) {
    for (j in 1:m) {
      samp[i,j,] <- rbeta(s, alpha[i, j], beta[i, j]) # samples from random beta distribution with alpha[i,j] beta[i,j]
    }
  }
  return samp
```

# Simulations

## Evaluation

### Mean Squared Error

When assessing the difference between two functions, we will frequently use the mean-squared error:

\begin{align*}
  MSE(\hat{y}, y) = \frac{1}{n}\sum_{i=1}^n (\hat{y}_i - y)^2
\end{align*}

where $\hat{y} \in \mathbb{R}^b$ is an $n$ element estimator for the function $y \in \mathbb{R}^n$.

### Frobenius Norm of Difference

When evaluating the similarity of two matrices, we will use the frobenius norm of the difference:

\begin{align*}
  ||X - Y||_F = \sqrt{\sum_{i=1}^n \sum_{j=1}^n |x_{ij} - y_{ij}|^2} = \sqrt{tr((X - Y)^*(X - Y))}
\end{align*}

## Convergence to True Distribution of Sampled Graphs

First, we will generate some simulation data using our $sample_beta_graph()$ function, and we will check whether the data is properly distributed. We will repeatedly use this function in this and later notebooks, so it will be critical for this function to work properly.


### Visualization

First, we will begin with a qualitative evaluation of the distribution of our 6 edges with respect to the true distribution with a small number of sampled graphs (100) and a large number of sampled graphs (1000). We expect that with a larger number of sampled graphs, we should get qualitatively better estimated distributions.

```{r fig.height=6, fig.width=8, warning=FALSE, message=FALSE}
require(wssg)
require(ggplot2)
require(reshape2)
library(MASS)
library(scales)
require(Rmisc)

lseq <- function(from, to, n) {
  return(round(exp(seq(log(from), log(to), length.out = n))))
}
mse <- function(x, y) {
  return(mean((x - y)^2))
}

xdim <- 2
ydim <- 3
alpha <- array(runif(6, min=.5, max=5), dim=c(xdim, ydim))  # true alpha
beta <- array(runif(6, min=.5, max=5), dim=c(xdim, ydim))  # true beta
n_vec <- c(100, 1000)
n_spaces <- length(n_vec)
n_bin <- 15
edge_err <- array(NaN, dim=c(xdim, ydim, n_spaces))  # the error per edge per iteration
err <- array(NaN, dim=c(n_spaces))  # the average error per iteration
hist_breaks <- seq(0, 1, length.out=n_bin)  # beta distr btwn 0 and 1

for (k in 1:length(n_vec)) {
  edge_distr_plots <- list()
  n <- n_vec[k]
  samp <- sample_beta_graph(alpha, beta, n)
  counter <- 1
  for (i in 1:xdim) {
    for (j in 1:ydim) {
      hsamp <- hist(samp[i, j,], breaks=hist_breaks, plot=FALSE)
      true <- dbeta(hsamp$mids, alpha[i, j], beta[i, j])
      edge_err[i, j, k] <- mse(hsamp$density, true)
      plot_data <- data.frame(x=hsamp$mids,
                              class=c(rep("yhat", length(hsamp$density)),
                                      rep("y", length(true))),
                              y=c(hsamp$density, true))
      edge_distr_plots[[counter]] <- ggplot(plot_data, aes(x=x, y=y, group=class, color=class)) +
        geom_line() +
        xlab('X') +
        ylab('Density') +
        ggtitle(sprintf('Edge (%d, %d), MSE=%.3f', i, j, edge_err[i, j, k])) +
        theme(text=element_text(size=10))
      counter <- counter + 1
    }
  }
  multiplot(plotlist=edge_distr_plots, layout=matrix(1:length(edge_distr_plots), nrow=xdim, byrow=TRUE))
}

```

### Simulation

For this simulation, we will do as follows:

+ Sample n observations from a 2x3 graph RV where each edge obeys the beta distribution with parameters alpha, beta.
+ Histogram bin the values into 20 bins from 0 to 1.
+ Show the average MSE (average over all of the edges) between the density estimation per edge and the true density function per edge approaches 0 as $n \rightarrow \infty$ as a function of n.
+ Show 100 number of samples histogram from a randomly selected edge, and show 1000 samples histogram from a randomly selected edge, with respect to the true distribution.

We know that by convergence, if we sample a large number of values our sample density should approximate to the true density better and better, and this simulation checks that this property is satisfied.

```{r }
n_vec <- lseq(50, 5000, 20)
n_spaces <- length(n_vec)
edge_err <- array(NaN, dim=c(xdim, ydim, n_spaces))  # the error per edge per iteration
err <- array(NaN, dim=c(n_spaces))  # the average error per iteration
for (k in 1:length(n_vec)) {
  n <- n_vec[k]
  samp <- sample_beta_graph(alpha, beta, n)
  for (i in 1:xdim) {
    for (j in 1:ydim) {
      hsamp <- hist(samp[i, j,], breaks=hist_breaks, plot=FALSE)
      edge_err[i, j, k] <- mse(hsamp$density, dbeta(hsamp$mids, alpha[i, j], beta[i, j]))
    }
  }
  err[k] <- mean(edge_err[,,k])
}

base_breaks <- function(n = 10){
    function(x) {
        axisTicks(log10(range(x, na.rm = TRUE)), log = TRUE, n = n)
    }
}

results <- data.frame(n=n_vec, error=err)
err_plot <- ggplot(results, aes(x=n, y=error)) +
  geom_point(size=3) +
  scale_x_log10(labels=trans_format("log10", math_format(10^.x)), breaks=trans_breaks("log10", function(x) 10^x, n=4)) +
  xlab('Number of Samples') +
  ylab('Average edge-wise MSE') +
  ggtitle('Analalyzing convergence of Samples to Truths') +
  theme(text=element_text(size=12))
err_plot + annotation_logticks(base = 10)
```

As we can see above, our edge-wise MSE drops off exponentially as the number of samples increases.


## Estimation is Consistent

Here, we will verify that our estimator converges for arbitrary random alpha, beta as our number of samples increases, given that our sampling code is correct (see previous section and test code). We are using the method of moments estimators for alpha and beta, and while the method of moments estimators might not necessarily converge as rapidly to the true parameters as the maximum likelihood estimators, they are still consistent.

For this simulation, we will do as follows:

+ Sample n observations from a 2x3 graph RV where each edge obeys the beta distribution with parameters alpha, beta.
+ Plot frobenius norm of difference between estimated alpha and true alpha (likewise for beta) as a function of the number of samples n.
+ Plot estimated alpha, beta for 100 samples and 1000 samples and compare qualitatively to the true alpha, beta.

Due to consistency, we expect that as n increases we should get closer and closer approximations of alpha, beta. 

```{r }

```
